The unphysicality of Hilbert spaces

### 2024-11-06 Articles/2023-HilbertQuantum/Submissions/Quantum Studies/HilbertQuantum.tex
HilbertQuantum: describing what states are
Further edits in response to reviewer comments

When we first drafted this paper, we naively thought that the advantage of having a well-posed mathematical structure that represents, and only represents, physically well-defined objects would be self-evident. It does not seem to be the case: a previous reviewer stated that ``the question of whether the limiting states are actually physical is rather pointless;'' conversely, others see no problem in working with QFT path integrals that are not mathematically well-defined. We are therefore in the absurd, from our perspective, position of having to defend that, yes, the math we need to use in physics has be mathematically precise and, yes, the math we need to use in physics has to represent, perhaps idealized, physically realizable objects. The cost of this failure is all around us: an ever increasing number of interpretations of quantum mechanics, abstract mathematical work whose connection to physics is unclear, theoretical physics more and more disconnected from experimental physics. The reader that, like us, finds this obvious can skip this section.

One problem is that most physicists see mathematics as just a tool for calculation, much like spectroscopes or telescopes are tools to gather data. As long as they do what they are supposed to do, the details are irrelevant because the physics lies somewhere else. If they do not work correctly, it is just a detail for the expert to fix. But mathematics plays a much more fundamental role in physics: it is the language in which we specify our models. When we say ``states are rays in a Hilbert space'' we are giving a physical specification. We are saying that there are objects in the real world that can be modeled with a particular mathematical structure. If Hilbert spaces do not work, it is not simply a math problem: the model is wrong, which is a physics problem. This is the question we are examining in this paper: does the math we use in quantum mechanics model the physical world correctly? Is our physical problem well specified?

To make this determination, one needs a standard to judge the physicality of a formal definition\cite{redei_tension_2020, north_physics_2021} and, rather tellingly, physics tradition does not provide one. Here is the standard we follow in our work.\cite{aop-book}  Suppose we are studying a physical system. A mathematical definition is \textbf{physical} if it properly characterizes the physical system in the given conditions. That is, if we are able to \textbf{justify on physical grounds} that the given mathematical definition is needed to capture and only capture a particular aspect of the physical system under the given condition. On the other hand, a mathematical definition is \textbf{unphysical} if it can be shown to require properties or operations that cannot have a physical counterpart. The question of this paper, then, is whether the use of Hilbert spaces to characterize quantum systems can be justified on physical grounds, from physical requirements. This is the task at hand.

While it seems obvious to us, it may be worth stating that \textbf{it is not up to mathematicians to justify the use of a particular mathematical structure in physics}. Ultimately, the justification cannot be solely in terms of mathematical matters, but rather it has to rely on that elusive ``physical intuition.'' In fact, the task is exactly to make that elusive ``physical intuition'' precise enough to turn it into mathematical definitions. This is the job of a physicist. In our specific case, the core argument we will use is that states described by measurable quantities with infinite or ill-defined expectation values are not physically realizable. To us, and to every experimental physicist we talked to, ``it doesn't make sense experimentally,'' and this is the best argument possible to show that something is unphysical. ``It can't be done.'' No argument is better. However, more than one previous reviewer found no problem with infinite (or undefined) expectation values, as they adopt an ``ensemble interpretation,'' and tried to convince us that we shouldn't either. How can progress on the mathematical foundations of quantum mechanics be made if an argument that is accepted by the vast majority of physicists is rejected?

If the problem of justifying the use of mathematical structures in physics is, ultimately, a physical problem, it also follows that it cannot be solved simply by a different mathematical approach. For example, according to the literature, (CITE) rigged Hilbert spaces were developed to give a mathematical foundation to Dirac’s bra-ket formalism, and be able to talk about eigenstates of position and momentum. The question is not whether these can be useful at a mathematical level to solve certain problem. They clearly are. The question is whether, on physical grounds, this is a good idea. Should we really treat continuous and discrete spectra the same, or realize that they are, on physical grounds, two rather different things? Should the fact that the energy spectrum is typically discrete for bound states and continuous for unbound states more seriously, and realize that there is something in the topology of the spectra that is rather physically important? Should we then make the different more prominent in our thinking, and not hide in mathematical details? Similarly, one can take the algebraic route and start with an algebra of observables. This, some argue, make some mathematical problems easier to solve. Again, no question about it. It still begs the question: why is it that, on physical grounds, observables should form a $C^*$-algebra? What, then, is the suitable physically meaningful state space? Moreover, note that in both those approaches Hilbert spaces still play a crucial role. If Hilbert spaces are physically problematic, these approaches are still relying on an object that is physically problematic.

While the argument is ultimately physical, and it is therefore the job of a physicist to make these arguments, it has to be a physicist that is well-versed in foundational issues in mathematics. The justification, in fact, will need precise understanding of the mathematical definitions and their implication for both the physics and the math. It is often the case that the mathematical detail informs the physics and shouldn't be swept under the rug.  Let us give an example. In a previous work we found that a set of physically distinguishable objects must be a $T_0$ second countable topological space where each open set represents a verifiable statement, a statement allowing a test that terminates in finite time if and only if the statement is true. This result tells us why functions in physics must be ``well-behaved'' (i.e.~topologically continuous in the right topology), why sets with cardinality greater than the continuum are unphysical (i.e.~they cannot be given a $T_0$ second countable topology) and why the Banach-Tarski paradox does not apply in physics (i.e. non-Borel sets are unphysical as they cannot be associated with an experimental test). This tells us that, at the very least, non-separable Hilbert spaces are unphysical as they have cardinality greater than the continuum.\cite{pittphilsci18363} This ties together ideas from point-set topology, logic, set theory, measure theory and computer science. Clearly, one needs to have at least a basic understanding of these fields to be able to follow the argument, and most degree programs in physics do not include typically discuss these topics in depth.\footnote{A previous reviewer conflated the completeness of the Hilbert space with the existence of a complete orthonormal basis of said space.} However, in the end, only a physicist can judge whether the original premise, the evidence-based requirement of physics, justifies the mathematical requirements.

To summarize, we are not interested in understanding whether Hilbert spaces have nice mathematical features for calculating or proving results. The fact that other tools have been developed to solve other mathematical problems is not relevant to this discussion. We are only interested in the question of mathematical representation of quantum states, which are physical entities we prepare in a lab. Something we produce, manipulate and, in the end, measure. What physical requirements are we capturing when we model them mathematically? Are Hilbert spaces capturing, and only capturing, physically tenable assumptions, or not?

>>>>

When we first drafted this paper, we naively thought that the advantage of having a well-posed mathematical structure that represents, and only represents, physically well-defined objects would be self-evident. It does not seem to be the case: a reviewer from a previous journal submission stated that ``the question of whether the limiting states are actually physical is rather pointless;'' conversely, others see no problem in working with quantum field theory path integrals that are not mathematically well-defined. We are therefore in the absurd, from our perspective, position of having to defend that, yes, the math we need to use in physics has be mathematically precise and, yes, the math we need to use in physics has to represent---perhaps idealized---physically realizable objects. The cost of this failure is all around us: an ever-increasing number of interpretations of quantum mechanics, abstract mathematical work whose connection to physics is unclear, theoretical physics more and more disconnected from experimental physics. The reader who, like us, finds this obvious can skip this section.

One problem is that most physicists see mathematics as just a tool for calculation, much like oscilloscopes or telescopes are tools to gather data. As long as they do what they are supposed to do, the details are irrelevant because the physics lies somewhere else. If they do not work correctly, it is just a detail for the expert to fix. But mathematics plays a much more fundamental role in physics: it is the language in which we specify our models. When we say, ``States are rays in a Hilbert space,'' we are giving a physical specification. We are saying that there are objects in the real world that can be modeled with a particular mathematical structure. If Hilbert spaces do not work, it is not simply a math problem: the model is wrong, which is a physics problem. This is the question we are examining in this paper: does the math we use in quantum mechanics model the physical world correctly? Is our physical problem well specified?

To make this determination, one needs a standard to judge the physicality of a formal definition\cite{redei_tension_2020, north_physics_2021} and, rather tellingly, physics tradition does not provide one. Here we describe the standard that we follow in our work.\cite{aop-book}  Suppose we are studying a physical system. A mathematical definition is \textbf{physical} if it properly characterizes the physical system in the given conditions. That is, if we are able to \textbf{justify on physical grounds} that the given mathematical definition is needed to capture and only capture a particular aspect of the physical system under the given conditions. On the other hand, a mathematical definition is \textbf{unphysical} if it can be shown to require properties or operations that cannot have a physical counterpart. The question of this paper, then, is whether the use of Hilbert spaces to characterize quantum systems can be justified on physical grounds, from physical requirements. This is the task at hand.

While it seems obvious to us, it may be worth stating that \textbf{it is not up to mathematicians to justify the use of a particular mathematical structure in physics}. Ultimately, the justification cannot be solely in terms of mathematical matters, but rather it has to rely on that elusive ``physical intuition.'' In fact, the task is exactly to make that elusive ``physical intuition'' precise enough to turn it into mathematical definitions. This is the job of a physicist. In our specific case, the core argument we will use is that states described by measurable quantities with infinite or ill-defined expectation values are not physically realizable. To us, and to every experimental physicist we talked to, ``It doesn't make sense experimentally,'' and this is the best argument possible to show that something is unphysical. ``It can't be done.'' No argument is better. However, more than one previous reviewer found no problem with infinite (or undefined) expectation values, as they adopt an ``ensemble interpretation,'' and tried to convince us that we should not be concerned either. How can progress on the mathematical foundations of quantum mechanics be made if an argument that is accepted by the vast majority of physicists is rejected when one tries to discuss it in the literature?

If the problem of justifying the use of mathematical structures in physics is, ultimately, a physical problem, it also follows that it cannot be solved simply by a different mathematical approach. For example, rigged Hilbert spaces were developed to give a mathematical foundation to Dirac’s bra-ket formalism\cite{Dirac1947}, and to be able to talk about eigenstates of position and momentum, working mathematically with both continuous and discrete bases on an interchangeable level\cite{Madrid_2005,Celeghini2016}. The question is not whether these can be useful at a mathematical level to solve certain problems. They clearly are. The question at hand is whether, on physical grounds, this is a good idea. Should we really treat continuous and discrete spectra the same, or should we consider that they are, on physical grounds, two rather different things? Should the fact that the energy spectrum is typically discrete for bound states and continuous for unbound states make us wonder if there is something in the differing topologies of the spectra that is physically important? Should we then make the difference more prominent in our thinking, rather than trying to construct mathematical tools that treat them the same?  Similarly, one can take the algebraic route and start with an algebra of observables. This, some argue, makes some mathematical problems easier to solve, which indeed it does. It still begs the question: why is it that, on physical grounds, observables should form a $C^*$-algebra? What, then, is the suitable physically meaningful state space? Moreover, note that in both these frameworks, Hilbert spaces still play a crucial role. If Hilbert spaces are physically problematic, these approaches are still relying on a mathematical structure that is physically problematic.

While the argument is ultimately physical, and it is therefore the job of a physicist to make these arguments, it has to be a physicist that is well-versed in foundational issues in mathematics. The justification, in fact, will need precise understanding of the mathematical definitions and their implications for both the physics and the math. It is often the case that the mathematical detail informs the physics and should not be swept under the rug.  Let us give an example. In a previous work we found that a set of physically distinguishable objects must be a $T_0$ second-countable topological space where each open set represents a verifiable statement, a statement allowing a test that terminates in finite time if and only if the statement is true\cite{aop-book}. This result tells us why functions in physics must be ``well-behaved'' (i.e.~topologically continuous in the right topology), why sets with cardinality greater than the continuum are unphysical (i.e.~they cannot be given a $T_0$ second-countable topology) and why the Banach-Tarski paradox does not apply in physics (i.e.~non-Borel sets are unphysical as they cannot be associated with an experimental test). This tells us that, at the very least, non-separable Hilbert spaces are unphysical as they have cardinality greater than the continuum.\cite{pittphilsci18363} This ties together ideas from point-set topology, logic, set theory, measure theory and computer science. Clearly, one needs to have at least a basic understanding of these fields to be able to follow the argument, and most degree programs in physics do not typically discuss these topics in depth.\footnote{A previous reviewer conflated the completeness of the Hilbert space with the existence of a complete orthonormal basis of said space.} However, in the end, only a physicist can judge whether the evidence-based requirement of physics justifies the mathematical requirements.

To summarize, we are not interested here in understanding whether Hilbert spaces have nice mathematical features for calculating or proving results. Moreover, the fact that other tools have been developed to solve other mathematical problems is not relevant to this discussion. We are only interested in the question of mathematical representation of quantum states, which are physical entities we prepare in a lab. States in physics are the output of a preparation procedure, a characterization that works for both ``pure states'' of mechanics as well as ensembles in statistical mechanics. They are what we produce, manipulate and, in the end, measure. What physical requirements are we capturing when we model them mathematically? Are Hilbert spaces capturing, and only capturing, physically tenable assumptions, or not?
